---
title: "Homework 3 Stat 433"
author: "Juliana Brandt"
date: "2/24/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(nycflights13)
library(ggplot2)
library(ggplot2)
library(lubridate)
library(tidyr)
```

## Exercise
In r4ds flights… What time of day should you fly if you want to avoid delays as much as possible? Does this choice depend on anything? Season? Weather? Airport? Airline? Find three patterns (“null results” are ok!). Write your results into Rmarkdown. Include a short introduction that summarizes the three results. Then, have a section for each finding. Support each finding with data summaries and visualizations. Include your code when necessary. This shouldn’t be long, but it might take some time to find the things you want to talk about and lay them out in an orderly way.

### Introduction

The following data is based on the three datasets, flights, weather, and airlines, merged into one. From analyzing the data, filtering out unecessary values, and using only important variables, I was able to determine three significant patterns. I looked at Carrier and Weather Stations verus the average flight delay. I then took that information and added the time of day to see how that would impact the findings. Next, I looked at Wind Speed compared to the Carrier and how that was affecting average delay. I then added the time variable once again to see if time of day affected this. Lastly, I looked at Months of the Year compared to Visibility and its effect on average delay. I then added the time of day once more to look at time of day verse these other variable. All results are below with a description.


## 1. Carrier, Weather Station (Origin), Average Delay, Time of Day
```{r include=FALSE}
w = weather %>% 
  select(time_hour, origin, wind_speed, precip, visib, temp)

x = flights %>% 
  mutate(total_delay = dep_delay + arr_delay, 
         hour.min = as.numeric(paste0(hour + minute/60))) %>% 
  inner_join(w, by = c("time_hour", "origin")) %>% 
  left_join(airlines, by = "carrier") %>% 
  group_by(hour.min, origin, carrier, name) %>% 
  summarise(avg_delay = mean(total_delay, na.rm = T)) %>%
  arrange(avg_delay)
```
Looking at the "Carrier compared to Weather Station (origin) and Average Delay" graph below, we can see ExpressJet Airlines Inc. tends to have delayed flights much more often than the rest of the airlines. We can also say the opposite of Southwest Airlines and Virgin America because they seem to be more on time. Unfortunately we don't have enough information to comment on Hawaiian Airlines and SkyWest Airlines.

Next we can see the "Carrier compared to Time of Day (hours) and Average Delay for every Weather Station (origin)" graph. The interesting part of this graph is that we can see that no matter the airline, the average delay increases throughout each day. This can be due to airlines getting behind schedule as the day progesses.
```{r echo=FALSE}

ggplot(x,aes(x=origin, y=avg_delay)) + 
  geom_point(color = "blue", alpha = 0.2, na.rm = T) + 
  facet_wrap(.~name, nrow=4) +
  scale_x_discrete(name="Origin - Weather Station") +
  scale_y_discrete(name="Average Delay") +
  labs(title = "Carrier compared to Weather Station (origin) and Average Delay")

ggplot(x,aes(x=hour.min, y=avg_delay)) + 
  geom_point(aes(color = origin), alpha = 0.2, na.rm = T) + 
  facet_wrap(.~name, nrow=4) +
  scale_x_continuous(name="Time of Day (hours)", limits = c(4,24)) +
  scale_y_discrete(name="Average Delay") +
  labs(title = "Carrier compared to Time of Day (hours) and Average Delay for every Weather Station (origin)")
```



## 2. Carrier, Wind Speed, Airline, Average Delay, Time of Day

Now, I am comparing wind speed against the normal variables average delay and time of day. Looking at the graph below titled "Carrier compared to Wind Speed and Average Delay", we can see some trends.There is a clear relationship with wind speed and number of flights. As wind speed increases, we can see that no matter the carrier the general trend is that there are less flights that take off. That being said, we can also flip that and look at the amount of flights taking off when the wind speed is low. There is a much larger cluster of flights taking off when there is less wind speed. However, looking at average delay, we don't see much of a trend. The delays are relatively consistent across the board. 

The next graph titled "Wind Speed in MPH compared to Time of Day and Average Delay" compares wind speed with time of day and how that relates to average delay. Again in this graph we can see that very few flights take off when there is a high wind speed. What I find interesting from this plot is that the no matter the wind speed, the middle of the day has the majority of the delayed flights, so that would most likely delay following flights throughout the day as well.
```{r include=FALSE}
y = flights %>% 
  mutate(total_delay = dep_delay + arr_delay, 
         hour.min = as.numeric(paste0(hour + minute/60))) %>%
  inner_join(w, by = c("time_hour", "origin")) %>% 
  left_join(airlines, by = "carrier") %>% 
  mutate(wind.int = as.numeric(signif(wind_speed,0))) %>% 
  group_by(hour.min, origin, wind_speed, wind.int, name) %>% 
  summarise(avg_delay = mean(total_delay)) %>%
  arrange(avg_delay)
```

```{r echo=FALSE}
ggplot(y,aes(x=wind_speed, y=avg_delay)) + 
  geom_point(aes(color = origin), alpha = 0.2, na.rm = T) + 
  facet_wrap(.~name, nrow=4) +
  scale_x_discrete(name="Wind Speed (MPH)") +
  scale_y_discrete(name="Average Delay") +
  labs(title = "Carrier compared to Wind Speed and Average Delay")

ggplot(y,aes(x=hour.min, y=avg_delay)) + 
  geom_point(color = "#00AFBB", na.rm = T) + 
  facet_wrap(.~wind.int, scales = "fixed", nrow=4) +
  scale_color_gradientn(colors = c("darkblue", "lightblue", "#00AFBB", "#FFDB6D")) +
  scale_x_continuous(name="Time of Day (hours)", limits = c(4,24)) +
  scale_y_discrete(name="Average Delay") +
  labs(title = "Wind Speed in MPH compared to Time of Day and Average Delay")
```


## 3. Months of the Year, Visibility, Temperature, Average Delay, Time of Day

Finally, we can compare months of the year, visibility, and temperature, with average delay and time of day. Looking at the graph titled "Months of the Year compared to Visibility and Average Delay", we can see many different trends. To start, when the visibility is low, there are more delays, and less flights taking off, which makes sense. Planes tend to not take off when there is low visibility. Part of this graph shows the temperature as well. We can see that in the colder months, December through March, the visibility is lower, and the flights are more likely delayed than those of the summer months, which also makes sense. From May to October, there is an upward trend. This means that as visibility increases to see further, the more the average delay. This can be attributed to an overflow of flights needing to leave or possibly more travel.

Lastly, I wanted to show these variables in a different way that included time of day as well. In the graph "Visibility in Miles compared to Time of Day and Average Delay", we can see that visibility has a possible relationship with time of day. Low visibility seems to occur more in the morning than in the afternoons. This could be attributed to the marine layer off the coasts, or clouds taking some time to clear up in the mornings before the sun comes out. We can also see in this graph that low visibility possibly has a correlation to a higher average delay. We can see that when the visibility is 0 miles, the average delay is high in comparison to other times of day and visibilities.

```{r include=FALSE}

z = flights %>% 
  mutate(total_delay = dep_delay + arr_delay, 
         hour.min = as.numeric(paste0(hour + minute/60))) %>%
  inner_join(w, by = c("time_hour", "origin")) %>% 
  left_join(airlines, by = "carrier") %>% 
  group_by(hour.min, visib, month, temp) %>% 
  summarise(avg_delay = mean(total_delay)) %>%
  arrange(avg_delay)
```

```{r echo=FALSE}
ggplot(z,aes(x=visib, y=avg_delay)) + 
  geom_point(aes(color = temp), 
                 alpha = 0.2, 
                 na.rm = T) + 
  scale_color_gradientn(colors = c("darkblue","blue", "lightblue","yellow","orange","red")) +
  facet_wrap(.~month, nrow=4) +
  scale_x_discrete(name="Visibility in Miles") +
  scale_y_discrete(name="Average Delay") +
  labs(title = "Months of the Year compared to Visibility and Average Delay")


ggplot(z,aes(x=hour.min, y=avg_delay)) + 
  geom_point(color = "#00AFBB", na.rm = T) + 
  facet_wrap(.~visib, scales = "fixed", nrow=4) +
  scale_color_gradientn(colors = c("darkblue", "lightblue", "#00AFBB", "#FFDB6D")) +
  scale_x_continuous(name="Time of Day (hours)", limits = c(4,24)) +
  scale_y_discrete(name="Average Delay") +
  labs(title = "Visibility in Miles compared to Time of Day and Average Delay")

```


